Informing IPCC accounting of forest carbon using the global forest carbon database (ForC v4.0)

### 2022-04-12 doc/manuscript/ESSDmanuscript.Rmd
work on C pools

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; Mg C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig. `r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).

@@

Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies. 
IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.

@@

### Biomass
Biomass includes living vegetation, above- and below-ground.

The IPCC defines aboveground biomass as "all biomass of living vegetation, both woody and herbaceous, above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].

Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006].

>>>>

The end goal of IPCC greenhouse gas inventories is to quantify greenhouse gas emissions to, or withdrawals from, the atmosphere on an annual basis, most commonly on a national level [@ipcc_2006_2006;@ipcc_2019_2019]. 
For each stratum of subdivision within a land-use category, annual stock changes ($\Delta C$; t C yr^-1^) are calculated as the sum of changes in various pools (section `r section_C_pools`), plus any harvested wood products.
For each pool, $\Delta C$ may be calculated using the "Gain-Loss Method", which takes the difference between gains and losses (influx and outflux variables in Fig. `r fig_variable_mapping`)<!--IPCC eq.2.4-->, or using the "Stock-Difference Method", which computes $\Delta C$ based on C stocks at two points in time<!--IPCC eq.2.5--> [@ipcc_2006_2006]. 
Thus, C cycle variables relevant to the IPCC methodology and to EFDB include C stocks, increments, and fluxes in the IPCC-defined pools (Fig. `r fig_variable_mapping`).

@@

Forest ecosystem C pools may be parsed in various ways, and while certain definitions and thresholds are more common than others, there is no single standard for measuring or reporting that is adhered to by all -- or even most -- studies. 
IPCC parses forest C pools into biomass (aboveground and belowground), dead organic matter (dead wood and litter), and soil organic matter (Table `r table_pools`).
While there is some flexibility around the components included in each pool, each national inventory must apply these in a consistent manner.
In this section, we define and review the IPCC definitions in the context of typical forest C estimation methodologies.

@@

### Biomass
Biomass includes living vegetation, above- and below-ground, both woody and herbaceous, but with a focus on woody plants and trees given their much greater potential to sequester large amounts of C [@ipcc_2006_2006].

Aboveground biomass, which is typically <200 t C ha^-1^ but can exceed 700 t C ha^-1^ [@anderson-teixeira_carbon_2021], is defined by the IPCC as "all biomass of living vegetation above the soil including stems, stumps, branches, bark, seeds, and foliage" [@ipcc_2006_2006].
IPCC's guidance is that the understory may be excluded the understory if it constitutes a "minor" component, *where quantitative definitions of "understory" and "minor" is not provided*<!--verify-->, but where a commonly applied minimum size sampling threshold for mature forests would be 10 cm stem diameter at breast height (DBH).
A recent study characterizing the contributions of trees in different DBH classes to ecosystem C stocks and fluxes found that trees 1 - 10 cm DBH contributed up to ~8% aboveground biomass, ~17% aboveground woody net primary productivity ($ANPP_{woody.stem}$), and ~20% woody mortality ($M_{woody}$) of mature closed-canopy forests worldwide [@piponiot_distribution_2022].
In regrowth forests, woodlands, or savannas, small trees and shrubs contribute a much larger proportion of C stocks and fluxes [@piponiot_distribution_2022; @refs], and, correspondingly, biomass estimates for these ecosystems tend include smaller size classes [e.g., @refs]. 
Beyond the minimum DBH sampled, forest censuses and biomass estimates also differ in their inclusion of life forms other than dicot trees -- including lianas, ferns, palms, and bamboo -- which in some places can reach large sizes and/or constitute a large fraction of forest C.
*[IPCC guidance]*
Further, it is important to note that this excludes standing dead wood, which is included in remote sensing biomass estimates [@duncanson_aboveground_2021].

Belowground biomass is defined as "all biomass of live roots" [@ipcc_2006_2006], a definition including both coarse roots, whose biomass is typically estimated based on stem censuses and allometries or belowground to aboveground biomass ratios, and fine roots, whose biomass is typically estimated via extraction of roots from soil samples.
The former, which is typically <40 t C ha^-1^ [@anderson-teixeira_carbon_2021], is methodologically linked to aboveground biomass estimates, sharing the same methodological sources of variation, but tending to be far more uncertain.
Fine root biomass generally constitutes a much smaller C pool [typically <5 t C ha^-1^, @anderson-teixeira_carbon_2021], and IPCC guidance is that it can be excluded when fine roots cannot be distinguished empirically from soil organic matter or litter [@ipcc_2006_2006], which can be a painstaking process.

### 2022-04-12 doc/manuscript/ESSDmanuscript.tex
work on C pools

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); Mg C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (section 2.1), plus any harvested wood products. For each pool,
\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
takes the difference between gains and losses (influx and outflux
variables in Fig. 1), or using the ``Stock-Difference Method'', which
computes \(\Delta C\) based on C stocks at two points in time
\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
methodology and to EFDB include C stocks, increments, and fluxes in the
IPCC-defined pools (Fig. 1).

@@

Forest ecosystem C pools may be parsed in various ways, and while
certain definitions and thresholds are more common than others, there is
no single standard for measuring or reporting that is adhered to by all
-- or even most -- studies. IPCC parses forest C pools into biomass
(aboveground and belowground), dead organic matter (dead wood and
litter), and soil organic matter (Table 1). While there is some
flexibility around the components included in each pool, each national
inventory must apply these in a consistent manner.

@@

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & acceptable to exclude understory if minor component\\
\hline
 &  & biomass allometry & \\
\hline
 &  & include dead standing? & no\\
\hline
belowground biomass & all biomass of live roots & min dbh (coarse roots) & \\
\hline
 &  & allometry or assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & \\
\hline
 &  & min root diameter & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
\hline
 &  & sampling depth & \\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & layers included & includes entire O horizon\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

Biomass includes living vegetation, above- and below-ground.

The IPCC defines aboveground biomass as ``all biomass of living
vegetation, both woody and herbaceous, above the soil including stems,
stumps, branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}.

Belowground biomass is defined as ``all biomass of live roots''
\citep{ipcc_2006_2006}.

>>>>

The end goal of IPCC greenhouse gas inventories is to quantify
greenhouse gas emissions to, or withdrawals from, the atmosphere on an
annual basis, most commonly on a national level
\citep{ipcc_2006_2006, ipcc_2019_2019}. For each stratum of subdivision
within a land-use category, annual stock changes (\(\Delta C\); t C
yr\textsuperscript{-1}) are calculated as the sum of changes in various
pools (section 2.1), plus any harvested wood products. For each pool,
\(\Delta C\) may be calculated using the ``Gain-Loss Method'', which
takes the difference between gains and losses (influx and outflux
variables in Fig. 1), or using the ``Stock-Difference Method'', which
computes \(\Delta C\) based on C stocks at two points in time
\citep{ipcc_2006_2006}. Thus, C cycle variables relevant to the IPCC
methodology and to EFDB include C stocks, increments, and fluxes in the
IPCC-defined pools (Fig. 1).

@@

Forest ecosystem C pools may be parsed in various ways, and while
certain definitions and thresholds are more common than others, there is
no single standard for measuring or reporting that is adhered to by all
-- or even most -- studies. IPCC parses forest C pools into biomass
(aboveground and belowground), dead organic matter (dead wood and
litter), and soil organic matter (Table 1). While there is some
flexibility around the components included in each pool, each national
inventory must apply these in a consistent manner. In this section, we
define and review the IPCC definitions in the context of typical forest
C estimation methodologies.

@@

\caption{\label{tab:table_pools}\textbf{IPCC-defined forest carbon pools with definitions and measurement methods.} Definitions from IPCC Table 1.1. (See Table 1.1 in IPCC guidance).}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\raggedright}X>{\raggedright}X>{\raggedright}X}
\hline
\textbf{pool} & \textbf{definition} & \textbf{important sources of estimate variation} & \textbf{IPCC guidance}\\
\hline
aboveground biomass & all biomass of living vegetation & min dbh & acceptable to exclude understory if minor component\\
\hline
 &  & include non-dicot trees & ?\\
\hline
 &  & include dead standing? & no\\
\hline
 &  & biomass allometry & ?\\
\hline
belowground biomass & all biomass of live roots & all factors relevant to aboveground biomass & see above\\
\hline
 &  & allometry or assumed ratio of belowground to aboveground biomass (IPCC table 4.4) & \\
\hline
 &  & min root diameter & fine roots may be excluded when they cannot be distinguished empirically from soil organic matter or litter\\
\hline
 &  & sampling depth & \\
\hline
dead wood & all non-living woody biomass not contained in the litter, either standing, lying on the ground, or in the soil & min diameter & 10 cm default, but may be chosen by country\\
\hline
 &  & include belowground? & \vphantom{1} yes\\
\hline
litter & all non-living biomass with a size greater than the limit for soil organic matter  and less than the minimum diameter chosen for dead wood, lying dead, in various states of decomposition above or within the mineral or organic soil & max diameter (= max diameter for deadwood) & 10 cm default, but may be chosen by country\\
\hline
 &  & layers included & includes entire O horizon\\
\hline
 &  & include belowground? & yes\\
\hline
soil organic matter & organic carbon in mineral soils to a specified depth & sampling depth & depth = 30 cm default, but may be chosen by country\\
\hline
\end{tabu}
\end{table}

@@

Biomass includes living vegetation, above- and below-ground, both woody
and herbaceous, but with a focus on woody plants and trees given their
much greater potential to sequester large amounts of C
\citep{ipcc_2006_2006}.

Aboveground biomass, which is typically \textless200 t C
ha\textsuperscript{-1} but can exceed 700 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is defined by the IPCC as ``all
biomass of living vegetation above the soil including stems, stumps,
branches, bark, seeds, and foliage'' \citep{ipcc_2006_2006}. IPCC's
guidance is that the understory may be excluded the understory if it
constitutes a ``minor'' component, \emph{where quantitative definitions
of ``understory'' and ``minor'' is not provided}, but where a commonly
applied minimum size sampling threshold for mature forests would be 10
cm stem diameter at breast height (DBH). A recent study characterizing
the contributions of trees in different DBH classes to ecosystem C
stocks and fluxes found that trees 1 - 10 cm DBH contributed up to
\textasciitilde8\% aboveground biomass, \textasciitilde17\% aboveground
woody net primary productivity (\(ANPP_{woody.stem}\)), and
\textasciitilde20\% woody mortality (\(M_{woody}\)) of mature
closed-canopy forests worldwide \citep{piponiot_distribution_2022}. In
regrowth forests, woodlands, or savannas, small trees and shrubs
contribute a much larger proportion of C stocks and fluxes
\citep{piponiot_distribution_2022, refs}, and, correspondingly, biomass
estimates for these ecosystems tend include smaller size classes
\citep[e.g.,][]{refs}. Beyond the minimum DBH sampled, forest censuses
and biomass estimates also differ in their inclusion of life forms other
than dicot trees -- including lianas, ferns, palms, and bamboo -- which
in some places can reach large sizes and/or constitute a large fraction
of forest C. \emph{{[}IPCC guidance{]}} Further, it is important to note
that this excludes standing dead wood, which is included in remote
sensing biomass estimates \citep{duncanson_aboveground_2021}.

Belowground biomass is defined as ``all biomass of live roots''
\citep{ipcc_2006_2006}, a definition including both coarse roots, whose
biomass is typically estimated based on stem censuses and allometries or
belowground to aboveground biomass ratios, and fine roots, whose biomass
is typically estimated via extraction of roots from soil samples. The
former, which is typically \textless40 t C ha\textsuperscript{-1}
\citep{anderson-teixeira_carbon_2021}, is methodologically linked to
aboveground biomass estimates, sharing the same methodological sources
of variation, but tending to be far more uncertain. Fine root biomass
generally constitutes a much smaller C pool \citep[typically \textless5
t C ha\textsuperscript{-1},][]{anderson-teixeira_carbon_2021}, and IPCC
guidance is that it can be excluded when fine roots cannot be
distinguished empirically from soil organic matter or litter
\citep{ipcc_2006_2006}, which can be a painstaking process.
